---
title: "Script_0"
output: html_document
---
 
# Introduction to R 

In this script we will learn the basics of working with R in RStudio. 
 
##  Working Directory

Your working directory is a location on your computer where R will look for your data files. You will need to tell R what your working directory is everytime you open RStudio, otherwise R will just guess. 

Let's see what R has currently set as your working directory


```{r}
getwd()
```

Check out the pathway that prints out in your console. This pathway is UNIQUE to your computer. Notice that R uses forward slash.

It is a best practice to create a folder for your R data and tell R to use that folder every time you open R using `setwd()`

Here's the pathway that I would like to use to set my working directory:    

`setwd("C:/Users/Samantha/Desktop/R Files")`

I have chosen to use a file on my desktop called "R Files" to keep my data. Can you edit my setwd() code to set your own working directory? 
Try it!


## Types of data in R

The most common type of data you will work with is something called a dataframe. This is very similar to an Excel spreadsheet in appearance.

There are two ways to create a dataframe in R, you can either
write one yourself or read in a pre-made data file. Most of the time
we will read in an external file. This tutorial will show you how to do
both. 

### A. Creating a dataframe 
Let's make one so you can see.

df is the name of the dataframe

```{r}
df <- data.frame(
  name = c("John", "Paul", "George", "Ringo"),
  birth = c(1940, 1942, 1943, 1940), 
  instrument = c("guitar", "bass", "guitar", "drums")
)
```


You can see what is in your dataframe with `View()`

```{r}
View(df)
```


Notice that we created 3 columns: name, birth, and instrument, then populated those columns with information inside the function `c()`.

------------------------
 Your turn! 
------------------------

Make your own dataframe that contains information on the band Aerosmith from the following data on instrument, birth year, and marital status for each member:

* Steven Tyler: lead vocals, born 1948, single       
* Joe Perry: guitar, born 1950, married        
* Tom Hamilton:  bass, 1951, married     
* Joey Kramer: drums, 1950, married    
* Brad Whitford: guitar, born 1952, married

```{r}
aerosmith <- data.frame(
  name = c("Steven Tyler", "Joe Perry", "Tom Hamilton", "Joey Kramer", "Brad Whitford"),
  birth = c(1948, 1950, 1951, 1950, 1952), 
  instrument = c("vocals", "guitar", "bass", "drums", "guitar"),
  status = c("single", "married", "married", "married", "married")
)
View(aerosmith)
```




### B. Reading in a data file 
The function for reading in a data file will differ depending on they type of file you are reading. In this example, we will use a csv

Here we are reading in the csv file, Philly_schools.csv

```{r}
schools <- read.csv("Philly_schools.csv", stringsAsFactors = FALSE)

View(schools)
```


This dataset contains information on public schools in Philadelphia. Each row represents 1 school. 


##  III. Classes of Variables


There are three main classifications for variables in R: 
* character    
* numeric    
* factor

### A. Character variables
 
Character variables are just text. In the schools dataframe, the SCHOOL_NAME_1 variable is a character- it is just the name of the school.

We can check to see if a variable is a character

```{r}
class(schools$SCHOOL_NAME_1)
```


Here we are asking R to give us the class type of the column for school name inside the dataframe, schools. This dataframe$columnname syntax will become quite familiar to you in time!


### B. Numeric Variables
 
Numeric variables are just numbers. The variables Average_salary is the average salary of teachers in each school. 

```{r}
class(schools$Average_salary)

```


### C. Factor Variables

Factors are a special type of variable. They will appear in a dataframe as either character or numeric, but they have some additional information associated with them that is hidden to us, but that R uses. Factor levels control how variables are graphed, displayed in tables, and matter for interpreting regressions. We'll discuss factors in greater detail later on. 

The school level variable is currently a character 

```{r}
class(schools$SCHOOL_LEVEL_NAME)
```


Change this variable to a factor variable

```{r}
schools$SCHOOL_LEVEL_NAME <- as.factor(schools$SCHOOL_LEVEL_NAME)
```


Check that this variable is now a factor

```{r}
class(schools$SCHOOL_LEVEL_NAME)
```


See the factor levels

```{r}
levels(schools$SCHOOL_LEVEL_NAME)
```


We'll see how R uses these levels later on in this script.


##  V. Basic Data Manipulation

To get us familiar with working in R, we will learn some basic functions commonly applied to numeric and character variables

### A. Basic math for numeric data


```{r}
mean(schools$Total_suspensions)
summary(schools$Total_suspensions)
```

Making a new variable with math

```{r}
schools$NewEnrollmentProp<- (schools$New_student / schools$Enrollment)*100
summary(schools$NewEnrollmentProp)
```

The table functions provides us a count of character variables


```{r}
table(schools$SCHOOL_LEVEL_NAME)
```


Aggregating data

```{r}
aggregate(schools$Total_suspensions, by=list(schools$SCHOOL_LEVEL_NAME), FUN=mean)
```


### B. Categorical data

Print an alphabetized list with sort()

```{r}
sort(schools$SCHOOL_NAME_1)
```



------------------------
Your turn!
------------------------

Find the average teacher salary in Philadelphia

```{r}

```


Find the average teacher salary in Philadelphia by school level

```{r}

```


## IV. Types of Variables

There are two main types of variables: categorical and continuous. 

Categorical variables are categories, like gender and race. 

Continuous variables are numeric variables that have an almost infinite number of possibilities, such as age, weight, and time

------------------------
 Your turn! 
------------------------

Look in the dataframe schools, and find two examples of categorical and continuous variables. 

```{r}

```



## VI. Graphing different Types of variables

Continuous and Categorical variables naturally lend themselves to  different types of graphs. 


Histrogram: Base R offers a frequency chart of numeric variables

```{r}
hist(schools$Low_income_family)
```


### A. Density plot displays similar information

```{r}
plot(density(schools$Low_income_family))
```


Boxplots a useful for understanding the spread of the data

```{r}
boxplot(schools$Total_suspensions ~ schools$SCHOOL_LEVEL_NAME)
```


Notice anything funny about this graph? What is causing it?


To change the levels:

```{r}
schools$SCHOOL_LEVEL_NAME<-factor(schools$SCHOOL_LEVEL_NAME, levels = c("ELEMENTARY SCHOOL", "MIDDLE SCHOOL", "HIGH SCHOOL", "CAREER AND TECHNICAL HIGH SCHL"))
levels(schools$SCHOOL_LEVEL_NAME)
```


Now everything looks right!

```{r}
boxplot(schools$Total_suspensions ~ schools$SCHOOL_LEVEL_NAME)
```



Scatter plot

```{r}
plot(schools$Low_income_family, schools$Gifted_education)

plot(schools$Attendance, schools$Gifted_education, col = schools$SCHOOL_LEVEL_NAME, pch=16)
legend(x=80, y=60, legend=levels(schools$SCHOOL_LEVEL_NAME), col=c(1:4), pch=16)

```

------------------------
 Your turn! 
------------------------

Find something interesting to you in the data and plot it!

```{r}

```



------------------------
 Simplifying Syntax
------------------------

The attach() function allows us to simplify our code by  removing the need to refer to the name of the dataframe 

 For example:

```{r}
attach(schools)

mean(Low_income_family)
```

Notice that the mean() function is working without the dataframe$column name syntax

------------------------
 Your turn!
------------------------

 Now you will analyze a new dataset using your new skills!

 1. Read in the csv file, "acs_orientation.csv"

```{r}
acs <- read.csv("acs_orientation.csv", stringsAsFactors = FALSE)
```

 This file contains information on the gender, years of education,
 race, and income of 500 residents of Philadelphia. 

 2. Find the number of males and females in this dataset

```{r}
table(acs$Sex)
```


 3. Find the average salary of this dataset

```{r}
mean(acs$Income)
summary(acs$Income)
```


 4. Find the average salary of this dataset by gender

```{r}
aggregate(acs$Income, by = list(acs$Sex), FUN = mean)
```


 5. Find the average salary of this dataset by race

```{r}
aggregate(acs$Income, by = list(acs$Race), FUN = mean)
```



 Plot the relationship between years of education and income

```{r}
plot(acs$Education, acs$Income, pch=16)
```







#################################################################
# Will add:

# prop.table 
# xtabs or table(x,y) depending


# Possible stuff to add:

# readstata13
# basic math/functions in 1st section?

