---
title: "Script_1"
output: html_document
---


Script # 1
Based on Chapter 2, 4, 6, 11, and 18 of r4ds

Learning objectives:  

1. Review basics of R      
2. workflow of an analysis         
3. Learn to how to "pipe"


You will learn to analyze data using R in this course. 
Even if you don't like to use R for the rest of your career, 
the data analysis principles that you will learn from this course can be 
applied to other software like Excel. Similarly, the coding skills and
habits that you will develop in this class are transferable to other languages 
like Phython. 

So this is a good time to develop good programming habits. These habits will 
reduce errors in your codes and enhance your quality of life.
Error messages in R are known to be harmful to our mental health,
because they are written by people who are native R speakers.
Compared to them, R will always be a second language for us, since we are 
are analysts. Our native language is public adminstration and public policy.


# Packages


It is a good habit to put all the packages listed at the front of the script. These are packages we will use in this script.

Here is the documentation for R packages: [https://www.rdocumentation.org/](https://www.rdocumentation.org/) 

Install packages: dplyr, nycfilights13, tidyverse, readr, and magrittr. Note that you need to combine the names into a vector with c().

```{r}
# install.packages(c("dplyr", "nycflights13", "tidyverse", "readr", "magrittr"))
```



Load the packages.

```{r}
library(dplyr) # A package that you will use to manage your data.
library(nycflights13) # A data set of flights coming and going from New York city.
library(tidyverse) # Official package that creates a alternative "universe" for R users.
library(readr) # Import data files that are saved in other formts like Excel.
library(magrittr) # The pipe, `%>%`, comes from the magrittr package by Stefan Milton Bache.
```


# Discussion 

Here is a style guide that makes your codes easier to understand and debug: [http://style.tidyverse.org/](http://style.tidyverse.org/).

Let's review them.

In fact, there are two packages that help you apply the style guide:

1. styler: allows you to interactively restyle selected text, files, or entire projects.    
2. lintr: performs automated checks to confirm that your conform to the style guide. 

## Welcome to "tidyverse"

You may notice the terms: "tidy" and "tidyverse" in the styple guide. 
In this class, you are learning a relatively new approach to coding in R.
Hadley Wickham, the chief scientist at R-studio, "invented" tidyverse.
You can learn more about him at http://hadley.nz/. 
His book R for Data Science is our main textbook in this class. 
And he developed many influential and popular packages in R that support
this alternative universe called "tidyverse." 

tidyverse approach to R is different than the traditional approach to R.
In the beginning I will show you both "Base-R" and tidyverse. Later we will
only use Base-R when it is more efficient for us. As you can tell, I am a 
ruthlessly pragmatic in my work. 

Ruthless pragmatism is defined as "accepting the current circumstances and making practical decisions based on them, at any cost." [https://www.quora.com/What-does-the-phrase-ruthlessly-pragmatic-mean](https://www.quora.com/What-does-the-phrase-ruthlessly-pragmatic-mean) 

The official website for tidyvese is: [https://www.tidyverse.org/](https://www.tidyverse.org/). 

## Review of basics

Here are a few basics of R that you already know. But we will review them to make sure that we are on the same page.

You can use R as a calculator.

```{r}
1 / 200 * 30
(59 + 73 + 2) / 3
sin(pi / 2)

```

You also know that R is an "object oriented" language. 
You can create new objects with `<-`
Structure of the code: <object_name> <- <value>.
Read that code as : "<object name> gets <value>." 
Keyboard shortcut in R studio for <- : `Alt + (minus sign)`.  

```{r}
x <- 3 * 4
```

You can inspect an object by typing its name:

```{r}
x
```


### Discussion 

Here is a collection of cheatsheets: [https://www.rstudio.com/resources/cheatsheets/](https://www.rstudio.com/resources/cheatsheets/) 

Download the cheatsheet for Base-R.    
Download the cheatsheet for rstudio-ide. 


### Everything starts with a name: 
Use snake_case for descriptive names. Object names must start with a letter, and can only contain letters, numbers, `_` and `.`. Use __snake_case__ where you separate lowercase words with `_`. 

Here are examples:
`i_use_snake_case`    
`otherPeopleUseCamelCase`    
`some.people.use.periods`     
`And_aFew.People_RENOUNCEconvention`

`this_is_a_really_long_name <- 2.5`

To inspect this object, try out RStudio's completion facility: type "this" in the Console, press TAB, add characters until you have a unique prefix, then press return.

## Calling functions

R has a large collection of built-in functions that you can use (or "call") like this: function_name(arg1 = val1, arg2 = val2, ...). Functions allow you to do things in R.

```{r}
seq(1, 10)
y <- seq(1, 10, length.out = 5)
y
```


You can save time by surrounding the assignment with parentheses, which will print out the object that you created:

(y <- seq(1, 10, length.out = 5))

## Practice

1. Press Alt + Shift + K. What happens?     
2. How can you get to the same place using the menus?

## Workflow: scripts

The most important keyboard shortcuts: Cmd/Ctrl + Enter. This executes the current R expression in the console. 

## Workflow: projects

Create projects to organize your work

Instruct RStudio not to preserve your workspace between sessions: This will force you to capture all important interactions in your code. 

There is a great pair of keyboard shortcuts that will work together to make sure you've captured the important parts of your code in the editor:

1. Press Cmd/Ctrl + Shift + F10 to restart RStudio.     
2. Press Cmd/Ctrl + Shift + S to rerun the current script.

## New way to organize your work

You already know the importance of the working directory. This is where R looks for files. RStudio shows your current working directory at the top of the console window.

# You can also print this out in R code by running `getwd()`:

```{r}
getwd()
```


### Discussion

Keeping your files organized may be the most important habit that you need to develop if you are going work with data. Many "experienced" analysts failed to properly develop this habit. This failure has trapped them into a unhappy life.

## Make your work self-contained and portable

Don't set your working directory using `setwd()`.
A blog describing the virtue of project-oriented workflow [https://www.tidyverse.org/articles/2017/12/workflow-vs-script/](https://www.tidyverse.org/articles/2017/12/workflow-vs-script/).

The author of the blog wrote:

> "The chance of the setwd() command having the desired effect -- making the file paths work -- for anyone besides its author is 0%. It's also unlikely to work for the author one or two years or computers from now."

You should use the project, which is "self-contained" and portable.

## RStudio projects

### Discussion 

R experts keep all the files associated with a project together.

## Practice

1. Create a directory; call it "Week1" using a file manager of your operating system.     
2. Put Script_1 and employee_salaries.csv into the directory into "Week1".    
3. Click File > New Project in R-studio.    
4. Select the option: existing directory.    
5. Now you should have a project called: Week1.    

Whenever you want to work on the documents related to Week1, start your work by picking the project "Week1". R-studio will keep track of all your work. You will not get errors because you can't find your data or objects.


## Pipes

### Discussion

We will be using "pipes" as a part of tidyverse.

Pipes are a tool for expressing a sequence of multiple operations. 

Introduction to magrittr. [https://cran.r-project.org/web/packages/magrittr/vignettes/magrittr.html](https://cran.r-project.org/web/packages/magrittr/vignettes/magrittr.html)

Examples of piping

Filtering observations from flights data

```{r}
not_cancelled <- flights %>% 
  filter(!is.na(dep_delay), !is.na(arr_delay))
```


Summarizing the time delays by year, month, and day

```{r}
not_cancelled %>% 
  group_by(year, month, day) %>% 
  summarise(mean = mean(dep_delay))

```


### Discussion 

Now time to do some programming

## Import data from Excel worksheet saved as cvs format

We will use readr to import data [https://cran.r-project.org/web/packages/readr/README.html](https://cran.r-project.org/web/packages/readr/README.html).

## Practice

Download the cheat sheet for Data Import at [https://www.rstudio.com/resources/cheatsheets/](https://www.rstudio.com/resources/cheatsheets/).

Import data

```{r}
employee_salaries <- read_csv("employee_salaries.csv")
```


I downloaded the data set "employee_salaries.csv" from Open Data Philly website:
[https://www.opendataphilly.org/dataset/employee-salaries-overtime](https://www.opendataphilly.org/dataset/employee-salaries-overtime). It contains the salaries of the City employees. You should examine this data set, not just because we will use this data in this course. If you are interested in working for the City of Philadelphia, you can learn the structure of the City work force including wages associated with particular jobs from this data. `read_csv()` can import spreadsheets saved as column seperated format.

## Practice

1. How many city employees in the data?      
2. How many employees the city of Philadelphia has in 2017?      
3. What is the highest salary of a city employee without overtime?      
4. What is the highest salary of a city employee with overtime?      
5. What is the job title of the highest paid city employee without overtime?      
6. What is the job title of the highest paid city employee with overtime?       


## Answers:


Before we start analyzing the data, we should examine it look at the data; make sure you loaded tidyverse. How many of these variables are numeric? Which ones are characters? 

```{r}
glimpse(employee_salaries)
View(employee_salaries)

```


# Do we have missing data?

```{r}
sum(is.na(employee_salaries$annual_salary))

```


Advanced example. You want to check missing data for all the variables in the data

```{r}
na_count <-sapply(employee_salaries, function(y) sum(length(which(is.na(y)))))
(na_count <- data.frame(na_count))

```


Note that there are 12 missing data for "title" or "annual_salary"

1. How many city employees in the data?

Base R

```{r}
length(employee_salaries$objectid)
```

With pipe and tidyverse

```{r}
employee_salaries %>% summarise(Count = n())

```


2. How many employees the city of Philadelphia has in 2017? Do you notice that the data is quarterly for each calendar year? This means if you are not careful, you will be counting the same people three times in 2017.

Base R

```{r}
table(employee_salaries$calendar_year,employee_salaries$quarter)
```

With pipe and tidyverse

```{r}
employee_salaries %>% 
  group_by(calendar_year, quarter) %>% 
  summarise(Count = n())

```

The correct answer is: 31833

3. What is the highest annual salary of a city employee without overtime?

Base R

```{r}
max(employee_salaries$annual_salary, na.rm = TRUE)

```


With pipe and tidyverse

```{r}
employee_salaries %>% 
  summarise(max(annual_salary, na.rm = TRUE))

```

4. What is the highest salary of a city employee including overtime? By computing answer to Question #3, we know that there is an employee with missing annual salary

Check the observation with missing annual_salary which(is.na(employee_salaries$annual_salary))

Confirm the fact that row number (134439) has missing value for annual salary

```{r}
data.frame(employee_salaries[134439,])
```

Replace missing value "NA" with 0

```{r}
employee_salaries$annual_salary[is.na(employee_salaries$annual_salary)] <- 0

```


Create a new variable total_salary

Base R

```{r}
employee_salaries$total_salary <- employee_salaries$annual_salary + employee_salaries$ytd_overtime_gross

```

With pipe and tidyverse

```{r}
employee_salaries <- employee_salaries %>% 
  mutate(total_salary = annual_salary + ytd_overtime_gross)

```


Answer the question

5. What is the job title of the highest paid city employee without overtime?

```{r}
employee_salaries$title[employee_salaries$annual_salary==max(employee_salaries$annual_salary)]

unique(employee_salaries$title[employee_salaries$annual_salary==max(employee_salaries$annual_salary)])
```


6. What is the job title of the highest paid city employee with overtime?

```{r}
employee_salaries$title[employee_salaries$total_salary==max(employee_salaries$total_salary)]

unique(employee_salaries$title[employee_salaries$total_salary==max(employee_salaries$total_salary)])

```



## Additional practice

1. Go to https://github.com/fivethirtyeight/data and explore the data sets.        
2. Pick a data set that you want to explore.      
3. Download the data set.       
4. Import it into R.       
5. Explore the data. 
